Formal Analysis of Rate-Monotonic Scheduling Implementation via Real-Time Maude

### 2015-09-26 RMS/submission.tex
writting section 4.5

(A3) All overhead, such as task swapping times, is ignored.

In this paper, our application of the RMS algorithm meets the
assumption~(A2), while we would try to relax (A1) and (A3) in our
analysis model.

\subsection{The Implementation of RMS}
The algorithm of the RMS is straightforward, and it is easy to
implement it in pratical systems. However, the way to implement it is
not unique. Our target implementation to analyze is part of a
realistic avionic operating system from our industrial
partner. Interrupts triggered by the clock would come every $T$, which
we call \emph{interrupt cycle}. When an interrupt request comes, if
the system is interruptable, i.e. the interrupt mask is cleared, the
funtion $schedule()$ will be called; otherwise $schedule()$ will be
called once the interrupt mask becomes cleared.  The pseudocode of
$schedule()$ is shown as Algorithm~\ref{a:schedule}, where $taskList$
is the list of periodic tasks, which are needed to be scheduled, and
assumed ordered descendingly by priorities, and where both variables
$taskList$ and $timer$ are global. In this implementation, there is
only one kind of interrupt, the period $T_i$ of each task is a
multiple of $T$, and the tasks are independent, meeting the
assumption~(A2).

\begin{algorithm}
  \caption{The C-Like Pseudocode of $schedule()$}
  \label{a:schedule}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts}
  \State \Call{$updateStatus$}{$taskList$};
  \State $timer = timer + 1$;
  \State $p = taskList$;
  \While{$p$} 
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return;
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts}
      \State $p\rightarrow function()$; \Comment{to run the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
	\State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\section{Formalizing the RMS Implementation}

>>>>

(A3) All overhead, such as task switching times, is ignored.

In this paper, we consider an implementation instead of the RMS
algorithm itself, thus the model would be more complicated than this
ideal formulation.  Assumptions~(A1) and (A3) will be relaxed to shape a more
realisitc analysis model, as shown below.

\subsection{The Implementation of RMS}
The algorithm of the RMS is straightforward, and it is easy to
implement it in pratical systems. However, the way to implement it is
not unique. Our target implementation to analyze is part of a
realistic avionic operating system from our industrial
partner. Interrupts triggered by the clock would come every $T$, which
we call \emph{interrupt cycle}. When an interrupt request comes, if
the system is interruptable, i.e. the interrupt mask is cleared, the
function $schedule()$ will be called; otherwise the handler function
$schedule()$ will be pending until the interrupt mask becomes cleared.
The pseudocode of $schedule()$ is shown as Algorithm~\ref{a:schedule},
where $taskList$ is the list of periodic tasks, which are needed to be
scheduled, and assumed ordered descendingly by priorities, and where
both variables $taskList$ and $timer$ are global. In this
implementation, there is only one kind of interrupt, the period $T_i$
of each task is a multiple of $T$, and the tasks are independent,
meeting assumption~(A2).

\begin{algorithm}
  \caption{The C-Like Pseudocode of $schedule()$}
  \label{a:schedule}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts}
  \State \Call{$updateStatus$}{$taskList$};
  \State $timer = timer + 1$; \label{l:timer}
  \State $p = taskList$;
  \While{$p$} 
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return;
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts}
      \State $p\rightarrow function()$; \Comment{to run the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
	\State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\section{Formalizing the RMS Implementation}
Based on the code, we model the implementation under the following
assumptions:

(A1') Computation time $C_i$ for each task $\tau_i$ is not constant,
varying randomly between minimum $C^{min}_i$ and maximum $C^{max}_i$.

(A2) Tasks are independent, such that they are ready to run at their
initiation times and can be preempted instanly.

(A3') The times for scheduling and task switching should be
considered, while other overhead is ignored.

These assumptions make our model different with the standard
formulation. For instance, due to the interrupt masking and
consideration of scheduling overhead, jobs corresponding to task
$\tau_i$ could not be initiated at exactly $kT_i$, but a bit later.
Note that (A2) says jobs are ready to run at their initiation times,
however, they cannot run at their initiation times, because scheduling
takes time.

\subsection{Some Basic Data Types}
In our model, tasks are identified with their indexes of sort
\verb|Nat| in the $taskList$. We define a sort \verb|MaybeNat|
wrapping \verb|Nat|s to refer to some task, with \emph{constructor}
\verb|some| followed by a \verb|Nat| $n$ indicating the task with
index $n$, and \verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}

A sort \verb|Stack| is introduced to simulate the stack of the system,
storing the tasks that are being interrupted, and equipped with
operations \verb|push|, \verb|pop| and \verb|peek| on it.
\begin{verbatim}
  op bottom : -> Stack [ctor] .
  op _#_ : Nat Stack -> Stack [ctor] .
\end{verbatim}

We also need a sort \verb|Interval| denoting the interval, within
which the computation time of a task varies, and a sort \verb|Counter|
to represent how long some task has run and how long it should run.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

At last, to make our model checkable by \emph{untimed model-checking},
it is reasonable to reset the global variable $timer$ when it reaches
an upperbound while increasing (see Line~\ref{l:timer} in
Algorithm~\ref{a:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upperbound would be the least common multiple of
the periods of all tasks.
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}

\subsection{Modeling the System States}
Abstracting from some details, the running system could be considered
as consisting of several parts: the tasks which are scheduled, the
scheduler itself, the hardware including registers and stacks, and the
interrupt source. The scheduler, i.e. the function $schedule()$, has
only the variable $timer$ to be interesting, thus it can be described
by a single variable. We will present the modeling of the other parts
one by one.

\subsubsection{Tasks}
In this paper, since we are trying to analyze the implementation of
the scheduler, each task will be abstracted from its functionality and
treated as a \verb|Counter|. As mentioned above, times for scheduling
and task switching will be also considered in the model. They are just
seen as two system tasks. Every task is modeled as an object instance
of some subclass of the base class \verb|Task|:
\begin{verbatim}
  class Task | cnt : Counter .
  op error : -> Object [ctor] .
\end{verbatim}
The attribute \verb|cnt| denotes the computation time of the task,
while \verb|error| is an object indicating some task which misses its
deadline.

A periodic task, which is needed to be scheduled, is an object
instance of the subclass \verb|PTask| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, period : NzNat, status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, just as
in the implementation:
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
As mentioned already, periodic tasks would be identified by their
indexes in the list.

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes:
\begin{verbatim}
  class SysTask .            subclass SysTask < Task .
\end{verbatim}
Different from periodic tasks, system tasks would be organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.
We abstract here from the detailed definitions thanks to the
similarity.

\subsubsection{Hardware} Our model considers two parts of hardware relating
to the interrupt handling inside the system, the registers and the
stack.

The set of registers is modeled as an object instance of the class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt masking flag and \verb|ir| for the
interrupt request flag.
\begin{verbatim}
  class Regs | pc : TaskID, mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| is a supersort of \verb|MaybeNat| and
\verb|Oid|, referring to some task.
\begin{verbatim}
  subsorts MaybeNat Oid < TaskID .
\end{verbatim}

Some operations, such as \verb|getPc|, \verb|setMask| and so on, are
also defined on the class \verb|Regs|.

Then the hardware is naturally described by the sort \verb|Hardware|:
\begin{verbatim}
  op [_;_] : Object Stack ~> Hardware [ctor] .
  mb ([ < O:Oid : Regs |> ; S:Stack ]) : Hardware .
\end{verbatim}

\subsubsection{Interrupt Source}
The interrupt source is modeled in our model as an object of class \verb|IntSrc|,
with attribute \verb|cycle| denoting the interrupt cycle:
\begin{verbatim}
  class IntSrc | val : Time, cycle : Time .
\end{verbatim}

\subsubsection{System}
The system in the model is just a composition of the parts introduced
above, with a sort \verb|System|\footnote{Some variable declarations
  are not shown for simplicity.}:
\begin{verbatim}
  op _____ : 
       TaskList Timer SysTasks Hardware Object ~> System [ctor] .
  mb (L T STS HW < O:Oid : IntSrc |>) : System .
\end{verbatim}

\subsection{Interrupt Requesting}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero| while time
advances. This is modeled by the following instanous conditional rule
applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals to \verb|zero|, and \verb|_.intReq| sets the
\verb|ir| flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
Periodic tasks are initiated sequently by exactly the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is modeled by
the recursive operation as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and \verb|update_with_| updating the status of individual task
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule})
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, status : ST > with TIMER
        = if ST == DORMANT then < O : PTask | status : READY >
          else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > with TIMER
       = if ST == RUNNING then < O : PTask | status : INTERRUPT >
         else < O : PTask |> fi [owise] .
\end{verbatim}
In the case where \verb|TIMER|($timer$) can be divided by the task's
period, i.e. the task should be initiated, if the status is
\verb|DORMANT|, it should be ready to run; otherwise it means the
previous job of the task is not finished, thus it misses its deadline,
producing an \verb|error|. In the other case where the tasks should
not be initiated, the status of the task changes only if it is
\verb|RUNNING|. We can easily see \verb|updateStatus_with_| behaves
the same as $updateStatus()$.

\subsection{Interrupt Handling}
\label{ss:inthandling}
When an interrupt request comes, it may not be ``seen'' immediately by
the system. It requires the \verb|mask| flag to be cleared. Once the
request is seen, it is handled by performing two stages: the interrupt
mechanism of the hardware (such as clearing \verb|ir|, pushing context
into stack and so on) and running the $schedule()$ function. This is
modeled by the following instanous rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The operation \verb|_.interrupt| models
the interrupt mechanism performed by the hardware and does four
things: (i)clearing the \verb|ir| flag, which means the request has
been handled; (ii)pushing the current \verb|pc| value into the stack,
storing the interrupted context; (iii)assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that $schedule()$ is
running; and (iv)setting the \verb|mask| flag, to mask the coming
interrupt requests.

It is worth pointing out that \emph{scheduling} in our model does not
refer to the stage in which $schedule()$ is running. We use it to
refer to the stage scheduling the first should-be-run periodic task,
i.e.  the code strictly before Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore,
\begin{definition}
  The \emph{scheduling time} talked about in this paper consists of
  the time for switching from a running task to the function
  $schedule()$ when handling interrupts, the time spent by
  $schedule()$ finding and setting the \emph{first} should-be-run
  periodic task, and the time for switching from $schedule()$ to that
  first should-be-run task.
\end{definition}
Similarly, \emph{switching} in our model refers to the stage finding
out the next should-be-run periodic task when one is finished.
\begin{definition}
  The \emph{switching time} in this paper consists of the time for
  switching from a finished task back into $schedule()$, the time
  spent by $schedule()$ finding and setting the \emph{next}
  should-be-run periodic task, and the time for switching from
  $schedule()$ to that next should-be-run task.
\end{definition}

\subsection{Timed Behaviours of the System}
